Dynamics of Laterally Propagating Flames in X-ray Bursts

### 2019-10-19 xrb/paper.tex
some comments

To understand the time and length scales involved in flame
propagation, we do a 1D simulation of a laminar flame using our
microphysics.  Figure~\ref{fig:flame} shows the flame properties for a 
% \AssignTo{Kiran and Mike will remake with the 10/10 boosted flame}
He flame using the standard conductivities and our simple reaction
network.  This flame had a density of $2\times 10^6~\gcc$ and 
% \MarginPar{double check}
temperature of $5\times 10^7$~K.  We observe that this flame speed is
about $10^5~\cms$, the flame width is about 40~cm, and it
takes about 3~ms to settle into a sustained flame.  Note that this speed
is quite small compared to the speed estimated in
\citet{spitkovsky2002}.  We expect a multidimensional flame to
accelerate due to hydrodynamics interactions (wrinkling, turbulence
interactions, directed flows feeding fuel into the flame, ...).

@@

Flame propagation rates from both the 1D laminar flames and full 2D simulations are given in Table \ref{table:flame_speeds}.

@@

We wish to create an initial atmosphere consisting of a hot
``post-flame'' region and a cooler atmosphere that the flame will
laterally propagate into.  We put the hot region at the very left of
the domain (the origin of the axisymmetric coordinates).  To create
these initial conditions, we produce two different hydrostatic models,
a ``hot'' model that will represent the perturbation that drives the
flame and a ``cool'' model that will represent the state ahead of the
flame.  These will have different scale heights.  To create these
models, we break the vertical structure of the atmosphere into 4
layers: the underlying star, a ramp-up to the base of the
accreted atmosphere, a fuel layer representing the bulk of the
atmosphere where the flame will propagate, and an outer, low density,
isothermal buffer above the atmosphere that allows us
to capture expansion and explosive dynamics.  
% This is sketched in
% Figure~\ref{fig:sketch}.

The temperature profile in the star and ramp region is given as:
\begin{equation}
T = T_\star + \frac{1}{2} (T_\mathrm{hi} - T_\star) \left [ 1 + \tanh\left( \frac{\tilde{y}}{2 \delta_\mathrm{atm}} \right ) \right ]
\end{equation}
with
\begin{equation}
\tilde{y} = y - H_\star - \frac{3}{2} \delta_\mathrm{atm}
\end{equation}
Here, $\delta_\mathrm{atm}$ is a characteristic width of the transition
ramp, $T_\mathrm{hi}$ is the highest temperature in the HSE model---it
will represent the base of the fuel layer.

@@

We create two models, a ``cool'' model representing the atmosphere
ahead of the flame and a ``hot'' (or perturbed) model that will
initiate the flame by driving the burning.  The parameters we use for
the model generation are listed in Table~\ref{table:params}.

We blend the hot and cold models laterally to produce the perturbation
needed to initiate a localized flame, with the hot model at the
origin of the axisymmetric geometry.  The blending is done as:
\begin{align}
p(x,y) &= f(x) p_\mathrm{hot}(y) + [1-f(x)] p_\mathrm{cool}(y) \\
\rho(x,y) &= f(x) \rho_\mathrm{hot}(y) + [1-f(x)] \rho_\mathrm{cool}(y) \\
X_k(x,y) &= f(x) {X_k}_\mathrm{hot}(y) + [1-f(x)] {X_k}_\mathrm{cool}(y)
\end{align}
with
\begin{equation}
f(x) = \begin{cases}
     1 & x < x_\mathrm{pert} \\
   1 - \frac{x - x_\mathrm{pert}}{\delta_\mathrm{blend}} & x_\mathrm{pert} \le x \le x_\mathrm{pert} + \delta_\mathrm{blend} \\
     0 & x > x_\mathrm{pert} + \delta_\mathrm{blend}
\end{cases}
\end{equation}
Since the equation of hydrostatic equilibrium is linear and our
blending is a linear combination of two models in hydrostatic
equilibrum, the blended model is also in vertical equilibrium initially.  We choose
$x_\mathrm{pert} = 1.024\times 10^4$~cm and $\delta_\mathrm{blend} = 2048$~cm.  Once
the blended model is constructed, we compute $T(x,y)$ and $(\rho e)(x,y)$
from the equation of state,
\begin{align}
  T(x,y) &= T(\rho(x,y), p(x,y), X_k(x,y)) \\
  (\rho e)(x,y) &= \rho(x, y) \cdot e(\rho(x,y), p(x,y), X_k(x,y)) 
\end{align}

@@

The initial model is created on a uniform grid at the resolution
corresponding to the finest level of refinement.  At regions of the
atmosphere that are not refined to the finest level, we interpolate
density, pressure, and composition on the grid and then obtain the
temperature from the EOS.  Because of the strong degeneracy at the
high densities at the bottom of the domain, the numerical error in the
interpolation can create a temperature jump there.\MarginPar{discuss better}

@@

% made with time_series.py in flame_wave_boost_10_10_paper with --skip 100
\begin{figure}[h]
\centering
\plotone{time_series}
\caption{\label{fig:time_series} Time series of the mean molecular weight of the 
flame for our standard 10/10 simulation.}
\end{figure}

Figure~\ref{fig:10_10_overview} shows the temperature, energy
generation rate, and $w$ component of the velocity (the out-of-plane
velocity induced by the Coriolis force).  This latter field
illustrates the hurricane effect set up by the laterally spreading
burning front.  In the energy generation rate plot, we see that the
burning is mostly concentrated toward the bottom of the layer, as
expected since the density is greatest there.  We see that the peak of
the burning has moved off of the symmetry axis, demonstrating that the
burning front is propagating to the right.  In the temperature plot,
we see the effect of our refinement criteria focusing only on the
atmosphere where we are most dense, with an artificial drop in the
temperature at the refinement boundary due to the interpolation
procedure of our initial model which focused on pressure and density
rather than temperature. \MarginPar{Remind the reader what this
  was/where they should be looking in the plot to see this temperature
  drop} As we'll see when the lower resolution case, we do not believe
that this affects the results.

@@

% made with overview.py
\begin{figure}[h]
\centering
\plotone{flame_wave_boost_10_10_overview}
\caption{\label{fig:10_10_overview} Temperature, energy generation rate, and out-of-plane velocity for the 10/10 simulation at 20~ms.}
\end{figure}

Our default resolution puts $\sim 80$ zones vertically in the
``cool'' \MarginPar{double check} model atmosphere height.  To
understand the effects of resolution, we also performed a run with one
fewer level, giving a $20~\mbox{cm}$ resolution overall (this is our
10/10-lores run).  Figure~\ref{fig:10_10_lowres} shows the fields at
20~ms.  The structure is largely the same as the 10/10 run, with
largely the same flame shape and position, and the same structure in
the energy generation rate.  Since we do not have a jump in refinement
right below the atmosphere, we don't see the temperature feature from
the initial model mapping there, but we do see some cooling in the
\isot{Ni}{56} region.  This is likely a resolution effect, again due
to the strong gradient in temperature at the base of the atmosphere.

% made with slice_multi_crop.py
\begin{figure}[h]
\centering
\plotone{flame_wave_boost_10_10_lowres_slice}
\caption{\label{fig:10_10_lowres} Temperature, mean molecular weight,
  energy generation rate, and out-of-plane velocity for the 10/10 low
  resolution simulation at 20~ms.}
\end{figure}

@@

\begin{figure}[h]
\centering
\plotone{flame_wave_boost_5_5_slice}
\caption{\label{fig:5_5_overview} Temperature, mean molecular weight, energy generation rate, and out-of-plane velocity for the 5/5 simulation at 20~ms.}
\end{figure}

@@

\begin{figure}[h]
\centering
\plotone{network_compare}
\caption{\label{fig:network} Mean molecular weight at 15~ms comparing a run with
  aprox13 vs a run with the iso7 network.}
\end{figure}

@@

To measure the propagation rate of the burning front, we track the
location beyond the hottest part of the flame where
$\dot{e}_\mathrm{nuc}$ first drops to less than $0.1 \%$ of
maximum. Using $0.1 \%$ of the peak value rather than the peak value
itself helps to mitigate fluctuations caused by localized fluid
motions, as the reduced value lies in a less turbulent part of the
burning region. We also average $\dot{e}_\mathrm{nuc}$ over the
vertical coordinate to further dampen these fluctuations. After an
initial transient period typically spanning $\sim 3$ ms, the flame
settles into a state of steady propagation. The position data here are
well fitted by a linear function of time (as seen in Figure
\ref{fig:flame_speed}), and the resulting slope gives the velocity of
the flame front.

@@

Figure~\ref{fig:abar_temp} shows the energy generation rate as a function of the
temperature and $\bar{A}$. \MarginPar{I'm assuming will be defined somewhere above}
The energy generation rate peaks at low $\bar{A}$, where there is a high fraction
of unburnt material. The burning raises the temperature of this material, such
that the peak temperature coincides with the peak energy generation rate. The
material then cools again as the burning converts the fuel into ashes,
increasing $\bar{A}$ and reducing the energy generation rate as there is
less available fuel. Along the base of the plot we see cool unburnt fluid and
ashes. In the center of the plot there are thin `trails' in phase space, which
could correspond to less common reaction pathways in the reaction network.

@@

In addition to extending our models to 3D and relaxing the approximations used in this study, in the future we plan to perform a number of additional further studies. These include investigating the effects of ignition latitude and different initial models. We also plan to model mixed H/He bursts. This will require a
different reaction network, and perhaps different (more relaxed)
resolution requirements. A long term goal is to include magnetohydrodynamics in our models. Although the magnetic fields of neutron stars exhibiting X-ray bursts are relatively weak, with $B \leq 10^8 - 10^9$G \citep{mukherjee2015magnetic} (at least compared to e.g. magnetars, which have $B\sim 10^{15}$G), it has been found by \citet{art-2016-cavecchi-etal} that even weak magnetic fields could have a non-negligible effect on the flame propagation, reducing confinement due to the Coriolis force and leading to increased flame speeds. Ultimately, we wish to link our simulations to observed light curves. To do this, we will need to explore radiation transport in order to model how the burst energy propagates through the outer layers of the neutron star atmosphere.

@@

We use a 2-d axisymmetric geometry, with a reflection boundary on the symmetry axis (left)
and Neumann boundaries on all other sides.  In this geometry, the solution behaves like
a spherical problem, and has the form:
\begin{equation}
  T(r) = T_1 + (T_2 - T_1) \left (\frac{t_0}{t + t_0} \right )^{3/2} e^{-r^2/(4 D (t + t_0))}
\end{equation}
where $t_0$ is a small time used to set the initial width of the
Gaussian, $r$ is the distance from the origin, and $D$ is the
diffusion coefficient.  $T_1$ is the ambient temperature and $T_2$ is
the temperature of the peak of the Gaussian.  We use $D = 1$, $T_1 = 1$, $T_2 = 2$,
and $t_0 = 0.001$.

@@

\begin{figure}[t]
\centering
\plotone{diffusion_temp_amr}
\caption{\label{fig:diffusion} Temperature for the diffusion test problem at $10^{-3}~\mbox{s}$
for a diffusion test with a $64\times 128$ zone base grid and one level of refinement.}
\end{figure}

>>>>

To understand the time and length scales involved in flame
propagation, we do a 1D simulation of a laminar flame using our
microphysics.  Figure~\ref{fig:flame} shows the flame properties for a 
% \AssignTo{Kiran and Mike will remake with the 10/10 boosted flame}
He flame using the standard conductivities and our simple reaction \MarginPar{standard?  this is a boosted flame, right?}
network.  This flame had a density of $2\times 10^6~\gcc$ and 
% \MarginPar{double check}
temperature of $5\times 10^7$~K.  We observe that this flame speed is
about $10^5~\cms$, the flame width is about 40~cm, and it
takes about 3~ms to settle into a sustained flame.  Note that this speed
is quite small compared to the speed estimated in
\citet{spitkovsky2002}.  We expect a multidimensional flame to
accelerate due to hydrodynamics interactions (wrinkling, turbulence
interactions, directed flows feeding fuel into the flame, ...).

@@

Flame propagation rates from both the 1D laminar flames and full 2D simulations are given in Table \ref{table:flame_speeds}.  \MarginPar{I think we should only report the laminar speeds here and defer the turbulent speeds until the section where we talk about how they are computed}

@@

We wish to create an initial atmosphere consisting of a hot
``post-flame'' region and a cooler atmosphere that the flame will
laterally propagate into.  We put the hot region at the very left of
the domain (the origin of the axisymmetric coordinates).  To create
these initial conditions, we produce two different hydrostatic models,
a ``hot'' model that will represent the perturbation that drives the
flame and a ``cool'' model that will represent the state ahead of the
flame.  These will have different scale heights.  To create these
models, we break the vertical structure of the atmosphere into 4
layers: the underlying neutron star, a ramp-up to the base of the
accreted atmosphere, a fuel layer representing the bulk of the
atmosphere where the flame will propagate, and an outer, low density,
isothermal buffer above the atmosphere that allows us
to capture expansion and explosive dynamics.  
% This is sketched in
% Figure~\ref{fig:sketch}.

The temperature profile in the star and ramp region is given as:
\begin{equation}
T = T_\star + \frac{1}{2} (T_\mathrm{hi} - T_\star) \left [ 1 + \tanh\left( \frac{\tilde{y}}{2 \delta_\mathrm{atm}} \right ) \right ]
\end{equation}
with
\begin{equation}
\tilde{y} = y - H_\star - \frac{3}{2} \delta_\mathrm{atm}
\end{equation}
Here, $\delta_\mathrm{atm}$ is a characteristic width of the
transition ramp, $T_\star$ is the temperature of the neutron star, and
$T_\mathrm{hi}$ is the highest temperature in the HSE model---it will
represent the base of the fuel layer.

@@

We create two models, a ``cool'' model representing the atmosphere
ahead of the flame and a ``hot'' (or perturbed) model that will
initiate the flame by driving the burning.  The parameters we use for
the model generation are listed in Table~\ref{table:params}. \MarginPar{reference and discuss the figure of initial model here}

We blend the hot and cold models laterally to produce the perturbation
needed to initiate a localized flame, with the hot model at the
origin of the axisymmetric geometry.  The blending is done as:
\begin{align}
p(x,y) &= f(x) p_\mathrm{hot}(y) + [1-f(x)] p_\mathrm{cool}(y) \\
\rho(x,y) &= f(x) \rho_\mathrm{hot}(y) + [1-f(x)] \rho_\mathrm{cool}(y) \\
X_k(x,y) &= f(x) {X_k}_\mathrm{hot}(y) + [1-f(x)] {X_k}_\mathrm{cool}(y)
\end{align}
with
\begin{equation}
f(x) = \begin{cases}
     1 & x < x_\mathrm{pert} \\
   1 - \frac{x - x_\mathrm{pert}}{\delta_\mathrm{blend}} & x_\mathrm{pert} \le x \le x_\mathrm{pert} + \delta_\mathrm{blend} \\
     0 & x > x_\mathrm{pert} + \delta_\mathrm{blend}
\end{cases}
\end{equation}
Since the equation of hydrostatic equilibrium is linear and our
blending is a linear combination of two models in hydrostatic
equilibrum, the blended model is also in vertical equilibrium initially.  We choose
$x_\mathrm{pert} = 1.024\times 10^4$~cm and $\delta_\mathrm{blend} = 2048$~cm.  Once \MarginPar{double check these numbers}
the blended model is constructed, we compute $T(x,y)$ and $(\rho e)(x,y)$
from the equation of state,
\begin{align}
  T(x,y) &= T(\rho(x,y), p(x,y), X_k(x,y)) \\
  (\rho e)(x,y) &= \rho(x, y) \cdot e(\rho(x,y), p(x,y), X_k(x,y)) 
\end{align}

@@

The initial model is created on a uniform grid at the resolution
corresponding to the finest level of refinement.  At regions of the
atmosphere that are not refined to the finest level, we interpolate
density, pressure, and composition on the grid and then obtain the
temperature from the EOS.

@@

% made with time_series.py in flame_wave_boost_10_10_paper with --skip 100
\begin{figure}[t]
\centering
\plotone{time_series}
\caption{\label{fig:time_series} Time series of the mean molecular weight of the 
flame for our standard 10/10 simulation.}
\end{figure}

Figure~\ref{fig:10_10_overview} shows the temperature, energy
generation rate, and $w$ component of the velocity (the out-of-plane
velocity induced by the Coriolis force).  This latter field
illustrates the hurricane effect set up by the laterally spreading
burning front.  In the energy generation rate plot, we see that the
burning is mostly concentrated toward the bottom of the layer, as
expected since the density is greatest there.  We see that the peak of
the burning has moved off of the symmetry axis, demonstrating that the
burning front is propagating to the right.  In the temperature plot,
we see the effect of our refinement criteria focusing only on the
atmosphere where we are most dense, with an artificial change in the
temperature at the refinement boundary due in part to the construction
of the initial model using the fine grid resolution for HSE. \MarginPar{Remind the reader what this
  was/where they should be looking in the plot to see this temperature
  drop} As we'll see when the lower resolution case, we do not believe
that this affects the results.

@@

% made with overview.py
\begin{figure}[t]
\centering
\plotone{flame_wave_boost_10_10_overview}
\caption{\label{fig:10_10_overview} Temperature, energy generation rate, and out-of-plane velocity for the 10/10 simulation at 20~ms.}
\end{figure}

Our default resolution puts $\sim 80$ zones vertically in the
``cool'' \MarginPar{double check} model atmosphere height.  To
understand the effects of resolution, we also performed a run with one
fewer level, giving a $20~\mbox{cm}$ resolution overall (this is our
10/10-lores run).  Figure~\ref{fig:10_10_lowres} shows the fields at
20~ms.  The structure is largely the same as the 10/10 run, with
largely the same flame shape and position, and the same structure in
the energy generation rate.  Since we do not have a jump in refinement
right below the atmosphere, we don't see the temperature feature from
the initial model mapping there, but we do see some cooling in the
\isot{Ni}{56} region.  This is likely a resolution effect, again due
to the strong gradient in temperature at the base of the atmosphere.
The strong agreement in the low resolution run to the 10/10 run gives
us confidence that we are capturing the flame physics properly.

% made with slice_multi_crop.py
\begin{figure}[t]
\centering
\plotone{flame_wave_boost_10_10_lowres_slice}
\caption{\label{fig:10_10_lowres} Temperature, mean molecular weight,
  energy generation rate, and out-of-plane velocity for the 10/10 low
  resolution simulation at 20~ms.}
\end{figure}

@@

\begin{figure}[t]
\centering
\plotone{flame_wave_boost_5_5_slice}
\caption{\label{fig:5_5_overview} Temperature, mean molecular weight, energy generation rate, and out-of-plane velocity for the 5/5 simulation at 20~ms.}
\end{figure}

@@

\begin{figure}[t]
\centering
\plotone{network_compare}
\caption{\label{fig:network} Mean molecular weight at 15~ms comparing a run with
  aprox13 vs a run with the iso7 network.}
\end{figure}

@@

To measure the propagation rate of the burning front, we track the
location beyond the hottest part of the flame \MarginPar{what does this mean?} where
$\dot{e}_\mathrm{nuc}$ first drops to less than $0.1 \%$ of
maximum. Using $0.1 \%$ of the peak value rather than the peak value
itself helps to mitigate fluctuations caused by localized fluid
motions, as the reduced value lies in a less turbulent part of the
burning region. We also average $\dot{e}_\mathrm{nuc}$ over the  \MarginPar{I think we should discuss the averaging first, before we discuss the peak value}
vertical coordinate to further dampen these fluctuations. After an
initial transient period typically spanning $\sim 3$ ms, the flame
settles into a state of steady propagation. The position data here are
well fitted by a linear function of time (as seen in Figure
\ref{fig:flame_speed}), and the resulting slope gives the velocity of
the flame front.

@@

Figure~\ref{fig:abar_temp} shows the energy generation rate as a function of the
temperature and $\bar{A}$. 
The energy generation rate peaks at low $\bar{A}$, where there is a high fraction
of unburnt material. The burning raises the temperature of this material, such
that the peak temperature coincides with the peak energy generation rate. The
material then cools again as the burning converts the fuel into ashes,
increasing $\bar{A}$ and reducing the energy generation rate as there is
less available fuel. Along the base of the plot we see cool unburnt fluid and
ashes. In the center of the plot there are thin `trails' in phase space, which
could correspond to less common reaction pathways in the reaction network.

@@

In addition to extending our models to 3D and relaxing the
approximations used in this study, in the future we plan to perform a
number of additional further studies. These include investigating the
effects of ignition latitude and different initial models. We also
plan to model mixed H/He bursts. This will require a different
reaction network, and perhaps different (more relaxed) resolution
requirements. A long term goal is to include magnetohydrodynamics in
our models. Although the magnetic fields of neutron stars exhibiting
X-ray bursts are relatively weak, with $B \leq 10^8 - 10^9$G
\citep{mukherjee2015magnetic} (at least compared to e.g. magnetars,
which have $B\sim 10^{15}$G), it has been found by
\citet{art-2016-cavecchi-etal} that even weak magnetic fields could
have a non-negligible effect on the flame propagation, reducing
confinement due to the Coriolis force and leading to increased flame
speeds. The effect of magnetic fields on the conductivities and
anisotropic conduction might also be important.  Ultimately, we wish
to link our simulations to observed light curves. To do this, we will
need to explore radiation transport in order to model how the burst
energy propagates through the outer layers of the neutron star
atmosphere.

@@

We use a 2-d axisymmetric geometry, with a reflection boundary on the symmetry axis (left)
and Neumann boundaries on all other sides.  In this geometry, the solution behaves like
a spherical problem, and has the form:
\begin{equation}
  T(r,t) = T_1 + (T_2 - T_1) \left (\frac{t_0}{t + t_0} \right )^{3/2} e^{-r^2/(4 D (t + t_0))}
\end{equation}
where $t_0$ is a small time used to set the initial width of the
Gaussian, $r$ is the distance from the origin, and $D$ is the
diffusion coefficient.  $T_1$ is the ambient temperature and $T_2$ is
the temperature of the peak of the Gaussian.  We use $D = 1$, $T_1 = 1$, $T_2 = 2$,
and $t_0 = 0.001$.

@@

\begin{figure}[t]
\centering
\epsscale{0.5}
\plotone{diffusion_temp_amr}
\caption{\label{fig:diffusion} Temperature for the diffusion test problem at $10^{-3}~\mbox{s}$
for a diffusion test with a $64\times 128$ zone base grid and one level of refinement.}
\epsscale{1.0}
\end{figure}
